Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2016-12-21 manuscript.tex
latest

\begin{figure}
\centering
\includegraphics[scale=0.8]{lottery.pdf}
\caption{\label{fig:lottery} Each iteration of our lottery model has three main elements. Each generation, propagules are produced by adults which are dispersed at random over the unoccupied territories. Lottery competition then occurs in each unoccupied territory: each genotype has a probability proportional to $b_i n_i c_i$ of securing the territory. Then occupied territories are freed up by adult mortality. In Eq. \eqref{eq:delttot} and most of the paper, only adults can die (red crosses), but we will also consider the case where juveniles die (blue cross; section ``Primary strategies and Grime's triangle'').} 
\end{figure}

Each iteration, adults produce $m_i$ new offspring (henceforth called ``propagules''). We assume adults cannot be ousted from occupied territories, so only propagules landing on unoccupied territories are included in $m_i$. Dispersal is random and unrestricted; each propagule has an equal probability of landing on any of the $U$ unnocupied territories. Loss of propagules during dispersal is subsumed into $m_i$ ($m_i$ only counts survivors which go on to contest the territory they land in), and we assume that propagules disperse independently of each other. In general, $m_i$ will increase with $n_i$, and will also depend on population density $N$. For example, if $b_i$ is the number of propagules produced per individual from genotype $i$, then $m_i=b_i(1-N/T)n_i$ would reflect the loss of propagules due to random dispersal to occupied territories. Nevertheless, for simplicity we will assume a linear function $m_i=b_i n_i$. Our justification is that the specific choice of $m_i$ makes little difference to our analysis here other than potentially complicating the mathematics. However, the choice of $m_i$ could matter in other applications, particularly when the mode of dispersal is a central concern [cite coexistence].

@@

The increase in $n_i$ over one iteration due to territorial acquisition, $\Delta_+ n_i$, is the sum of genotype $i$'s victories over all $U$ unoccupied territories. Since $p_1(x_1)\ldots p_G(x_G)$ is equal to the proportion of unoccupied territories with $x_1,\ldots,x_G$ of the respective propagules (where $G$ is the number of genotypes present; and again, we assume that $T$ is large enough that fluctuations in this proportion are negligible), this sum can be replaced by an expectation over the $p_i$. This gives
\begin{equation}
\Delta_+ n_i(t)=U(t)\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G). \label{eq:growthsumuncoupled}
\end{equation}

@@

The propagule numbers $x_i$ in different territories are not independent random variables. To determine the dispersal outcomes in all unoccupied territories exactly, we would need to proceed territory-by-territory as follows. In the first territory we evaluate, $x_i$ drawn from a binomial distribution with $m_i$ trials and success probability $1/U$. In the second, $x_i$ is drawn from a binomial distribution with $m_i-x$ trials and success probability $1/(U-1)$, where $x$ is the number of propagules that landed in the first territory. And so on.

>>>>

\begin{figure}
\centering
\includegraphics[scale=0.8]{lottery.pdf}
\caption{\label{fig:lottery} Each iteration of our lottery model has three main elements. First, propagules are produced by adults which are dispersed at random over the unoccupied territories (only propagules landing on unoccupied territories are shown). Lottery competition then occurs in each unoccupied territory: each genotype has a probability proportional to $b_i n_i c_i$ of securing the territory. Then occupied territories are freed up by adult mortality. In Eq. \eqref{eq:delttot} and most of the paper, only adults can die (red crosses), but we will also consider the case where juveniles die (blue cross; section ``Primary strategies and Grime's triangle'').}
\end{figure}

Each iteration, adults produce $m_i$ new offspring (henceforth called ``propagules''). We assume adults cannot be ousted from occupied territories, so only propagules landing on unoccupied territories are included in $m_i$. Dispersal is random and unrestricted; each propagule has an equal probability of landing on any of the $U$ unnocupied territories. Loss of propagules during dispersal is subsumed into $m_i$ ($m_i$ only counts survivors which go on to contest the territory they land in), and we assume that propagules disperse independently of each other. In general, $m_i$ will increase with $n_i$, and will also depend on population density $N$. For example, if $b_i$ is the number of propagules produced per individual from genotype $i$, then $m_i=b_i(1-N/T)n_i$ would reflect the loss of propagules that disperse to occupied territories. Nevertheless, we will assume $m_i=b_i n_i$, meaning that all propagules land on unoccupied territories (a form of directed dispersal). This choice simplifies the mathematics without seriously restricting the generality of our analysis, since the results presented here are not sensitive to the specific functional form of $m_i$.

@@

The increase in $n_i$ over one iteration due to territorial acquisition, $\Delta_+ n_i$, is the sum of genotype $i$'s victories over all $U$ unoccupied territories. The expected proportion of unoccupied territories with $x_1,\ldots,x_G$ of the respective propagules is $p_1(x_1)\ldots p_G(x_G)$ ($G$ is the number of genotypes present), and the probability of $i$ winning in each of these territories is $c_i x_i/\sum_j c_j x_j$. As above, we assume that $T$ and the $n_i$ are large enough that we can ignore fluctuations in the proportion of unoccupied territories as well as the number of victories, which implies
\begin{equation}
\Delta_+ n_i(t)=U(t)\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G), \label{eq:growthsumuncoupled}
\end{equation}
where the sum only includes territories with at least one propagule present.

@@

The propagule numbers $x_i$ in different territories obey a multinomial distribution with

are not independent random variables.
